Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-18 submission/manuscript.Rmd
update germline gene text

Intriguingly, many _Kdm5c_-KO testis-enriched DEGs have functions specific to male germ cells, such as _Cytochrome C, testis (Cyct)_, a (Figure 1C), a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002]. We therefore wondered if the high enrichment of testis genes within the _Kdm5c_-KO brain reflected a failure to demarcate the soma and germline, as the testis contains both germ cells (e.g. spermatogonia) as well as somatic cells (e.g. Leydig cells) that support hormone production and germline functions. Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including cellular process involved in reproduction in multicellular organism (GO:0022412), spermatid development (GO:0007286), and sperm axoneme assembly (GO:0007288) (Figure 1A).

To validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B). We additionally assessed testis DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C). Altogether, these data suggest testis genes expressed within the _Kdm5c_-KO brain are actually male germ cell genes.

>>>>

Intriguingly, many _Kdm5c_-KO testis-enriched DEGs have functions specific to male germ cells, such as _Cytochrome C, testis (Cyct)_, a (Figure 1C), a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002]. We therefore wondered if the high enrichment of testis genes within the _Kdm5c_-KO brain reflected a failure to demarcate the soma and germline, as the testis contains both germ cells (e.g. spermatogonia) as well as somatic cells (e.g. Leydig cells) that support hormone production and germline functions. Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including spermatid development (GO:0007286) and sperm axoneme assembly (GO:0007288) (Figure 1A).

To validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B). The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C). Altogether, these data suggest testis genes expressed within the _Kdm5c_-KO brain are actually male germ cell genes.
